Integer Programming Formulations for the Shared Multicast Tree Problem

### 2018-09-26 smt/response.tex

\section*{Reviewer 3}
Our response in \textcolor{blue}{blue} follows the corresponding remark from the reviewer in \textcolor{red}{red}.

\begin{itemize}
\item \textcolor{red}{The paper provides two new ILP formulations for the Shared Multicast Tree (SMT) problem in wireless ad hoc networks.
To address the problem of ILP scalability, the paper also provides a study and proofs of related theorems that shed light on the quality of
solutions achievable by  LP  bounds.
As the number and also provides a new constraint generation scheme.
Numerical results demonstrate the usefulness of the approach. The paper is well written.} \textcolor{blue}{
We are grateful to the reviewer for her/his encouraging remarks, and we appreciate the (s)he finds the paper well written.
}
\item \textcolor{red}{Overall, the methodology makes contribution in the field of optimization as the approach can be used to address other similar problems
that can be modeled as ILP which is not scalable.
However, this paper aims to address the problem of SMT in wireless ad hoc networks and it is
really motivated by the important problem of saving energy in sensor networks.
There I see a certain minor weakness.
The model considered ignores details of energy consumption is sensor network
(see for example Halgamuge et al., Progress in Electromagnetics Research, 12, 2009, or other sources).
It will be good to my opinion to point out this weakness and provide certain qualitative comments to benefit the practitioners,
on how the work can be used in practice, or what is still needed to be done so the work can be useful.} \textcolor{blue}{
We thank the reviewer for suggesting this very useful reference, a citation of which has been included in the revised version of the paper.
We fully agree that there are gaps between the coarse energy model applied in our work and more detailed energy models for energy consumption is sensor networks,
e.g.\ those analyzed by Halgamuge et al. (2009).
These gaps, and their possible implications on practical usability of our approaches, are now discussed in the new Section 3.2.
}
\end{itemize}

\noindent
Final comment: 
Newly added parts of the paper are also in \textcolor{blue}{blue} font.
In cases when the text is restructured, but the sentences and expressions are not altered, the font color remains black.
Such updates should be apparent due to new highlighted sections.

### 2018-09-26 smt/sec7-heur.tex
MI (SMT): Minor changes in several files, mainly in Sect 8 - description of the results.

\textcolor{blue}{
To the best of our knowledge, the scientific literature offers few heuristic methods for SMT besides those studied by \citet{ivanova16isco}.
However, recent progress by \citet{pajor18} on heuristic methods for the related minimum Steiner tree problem can smoothly be exploited in order to generate analogous methods for SMT.
The main algorithmic components in the most basic heuristic are a construction algorithm, a local improvement algorithm,
and an algorithm for merging two Steiner trees in order to create a better one.
Construction and local improvement are applied in order to generate Steiner trees of presumably low cost.
If the cost is sufficiently low, the tree is stored in an ordered set $\mathcal{T}$ of at most $\gamma$ distinct trees.
Diversity among the trees is achieved by perturbing the edge costs randomly before construction and local improvement are applied.
After each tree generation, the new tree is merged with a randomly chosen tree from $\mathcal{T}$, and the resulting tree is kept in $\mathcal{T}$ if its cost is small.}

>>>>

\textcolor{blue}{
To the best of our knowledge, the scientific literature offers few heuristic methods for SMT besides those studied by \citet{ivanova16isco}.
However, recent progress by \citet{pajor18} on heuristic methods for the related minimum Steiner tree problem can smoothly be exploited in order to generate analogous methods for SMT.
The main algorithmic components in the most basic heuristic are a construction algorithm, a local improvement algorithm,
and an algorithm for merging two Steiner trees in order to create a better one.
Construction and local improvement are applied in order to generate Steiner trees of presumably low cost.
If the cost is sufficiently low, the tree is stored in an ordered set $\mathcal{T}$ of at most $\gamma$ distinct trees.
Diversity among the trees is achieved by perturbing the edge costs randomly before construction and local improvement are applied.
After each tree generation, the new tree is merged with a randomly chosen tree from $\mathcal{T}$, and the resulting tree is kept in $\mathcal{T}$ if its cost is small.
%After each tree generation, the new tree is merged with a randomly chosen tree from $\mathcal{T}$, 
%and the resulting tree is kept in $\mathcal{T}$ if its cost is smaller than some other tree already in $\mathcal{T}$.
%If there is no empty capacity in $\mathcal{T}$, its most expensive tree is discarded.
}

### 2018-09-26 smt/sec8-exp.tex
MI (SMT): Minor changes in several files, mainly in Sect 8 - description of the results.

The practical part of this work focuses on comparison of models presented in Sections \ref{sec:ILP} and \ref{sec:strong}.
Instances of given number of nodes and destinations are generated with random coordinates uniformly distributed on the square with corners at [0, 0] and [100, 100].
A time limit of 20 CPU-minutes is imposed on each run.
All experiments are run on an Intel Core 2 Quad CPU at 2.83 GHz and 8 GB RAM.
The models are implemented in Java with Concert technology and solved using CPLEX 12.5.1 optimizer.

@@

Small instances are often solved to optimality relatively quickly with B\&B.
We investigate what is the maximum size of instances that can be solved to optimality within 20 minutes using models $\mathcal{X}_1$ and $\mathcal{F}_1$.
The experiments compare sets of instances with different $|V|/|D|$ ratios.
For instances of the size given in columns 1--2, Table \ref{tab:small_inst} shows the average running time in CPU-seconds (columns 3--4),
the number of instances solved to optimality (columns 5--6),
and the average ratio between the upper and lower bounds on the minimum cost (columns 7--8) upon interruption of the solver.
Each table value is obtained after solving 25 instances of given size.

@@

\textcolor{blue}{\subsection{Comparison with alternative methods}}

\textcolor{blue}{
To the best of our knowledge, SMT has not been studied extensively in terms of inexact methods.
However, there is a manifold of a thorough research of the minimum Steiner tree problem, with the possibility of adapting the methods on SMT.
We turn our attention to recently published paper \citet{pajor18} which contains a metaheuristic algorithm for a fast generation of good solution to the Steiner tree problem.
The method is general, and is therefore very suitable for adaptation to our purposes.  The high level description is as follows:}

\textcolor{blue}{The construction procedure and local search is the adapted BIP algorithm together with refinements presented in \citet{ivanova16isco}.
The remaining parts of the algorithm are carried out as discussed in \citet{pajor18}.
The randomization is achieved by perturbing the edge costs in the input graph.
The obtained solution $S$ is evaluated with respect to the real (original) edge costs and added to the pool $P$ if the pool capacity is not filled.
If the pool is full and its worst solution $\hat{S}$ is worse than $S$, $\hat{S}$ is replaced by $S$ in $P$.
Then a new solution $\tilde{S}$ is constructed by combining some solution in $P$ with $S$. 
First, we adjust edge costs resulting in a modified input instance, 
and then run the construction and local search yielding $\tilde{S}$.
This new solution is again added to the pool if it is better than some solution already in $P$.
The adjustment of the edges tries to promote edges that appear in both $S$ and a selected solution in $P$. 
In that case, the original cost is preserved.
If an edge is present in only one of the two solutions, its cost is multiplied by a random integer between 100 and 500.
If an edge appears in neither of the solutions, its edge is multiplied by 1000, which makes it very unlikely to appear in the result $\tilde{S}$ of the combination.
}
\begin{table}[]
\centering
\setlength{\tabcolsep}{6pt} % Default value: 6pt
\renewcommand{\arraystretch}{1.4} % Default value: 1
\textcolor{blue}{
\begin{tabular}{rrrrrrrrrrr}
\multicolumn{1}{l}{}      & \multicolumn{1}{l}{}      & \multicolumn{3}{c}{5 min}  & \multicolumn{3}{c}{10 min} & \multicolumn{3}{c}{20 min}     \\\hline
\multicolumn{1}{l}{$|V|$} & \multicolumn{1}{l}{$|D|$} & \multicolumn{1}{c}{$\mathcal{X}_1$} & \multicolumn{1}{c}{$\mathcal{F}_1$} & \multicolumn{1}{c}{$\mathcal{M}$} & \multicolumn{1}{c}{$\mathcal{X}_1$} & \multicolumn{1}{c}{$\mathcal{F}_1$} & \multicolumn{1}{c}{$\mathcal{M}$} & \multicolumn{1}{c}{$\mathcal{X}_1$} & \multicolumn{1}{c}{$\mathcal{F}_1$} & \multicolumn{1}{c}{$\mathcal{M}$} \\
\hline
18  & 12   &100.0&100.0&101.8&100.0&100.0&100.0&100.0&100.0&100.0\\
21  & 14   &106.5&107.4&101.2&102.7&103.1&100.1&100.9&100.3&100.0\\
24  & 16   &144.2&112.2&101.3&124.0&114.0&100.1&104.3&112.7&100.0\\     
\hline
20  & 10   &100.6&100.0&101.4&100.1&100.0&100.0&100.1&100.0&100.0\\
22  & 11   &102.1&103.2&101.7&101.3&100.7&100.1&100.7&100.2&100.1\\
24  & 12   &106.6&111.7&101.7&105.0&105.5&100.0&103.1&106.2&100.0\\                               
\hline
21  & 7    &100.1&100.0&100.0&100.0&100.0&100.0&100.0&100.0&100.0\\
24  & 8    &102.3&101.4&102.0&102.1&100.7&100.1&101.1&100.1&100.0\\
27  & 9    &127.4&114.0&101.6&111.2&109.0&100.1&108.2&104.1&100.0                                  
\end{tabular}
}
\caption{\textcolor{blue}{Comparison of the metaheuristic algorithm and solving IP models with imposed time-limit.}}
\label{tab:metamod}
\end{table}

\textcolor{blue}{
Tab.\ref{tab:metamod} summarizes the comparison of the metaheuristic algorithm with the ability of IP models to produce feasible solutions when restricted time limit.
Each entry represents the average percentage  of the objective value yielded by the most successful method in a given instance.
For example, if all methods find the optimum in every instance, the three corresponding values are all 100.0.
Thus, the lowest value in a given instance type indicates that the corresponding method performed the most successfully. 
The averages are calculated over all instances of given $V$ and $D$.
Experiments with results in Tab. \ref{tab:small_inst} and Tab. \ref{tab:metamod} are carried out on the same sets of instances.
}

\textcolor{blue}{
The resulting values reveal that the metaheuristic algorithm converges quickly, and is often able to find optimal solutions faster than solving the IP models.
The solutions are, however, not guaranteed to be optimal.
The optimality is proven only on smaller instances, where the B\&B finds the optimum before the time limit is exceeded.
It is likely that the inexact method run on large instances does not find the optimum, but it very clearly outperforms B\&B, 
where the solutions found within the given time limit are far from the optimum.
}

@@

As the instance size increases, exact solution of the SMT problem becomes unrealistic.
To assess the quality of fast and possible inexact methods \citep{ivanova16isco}, access to tightest possible lower bounds is crucial.
In the next set of experiments, we compare two different methods for computing the bounds.
Section \ref{sec:expcomplp} indicates that $z\left(\text{LP}(\mathcal{X}_3)\right)$ is a strong bound,
but solving $\text{LP}(\mathcal{X}_3)$ requires excessive computations.
At this point, the CG scheme (Alg.\ \ref{alg:cg}) comes into play.
Because $\text{LP}(\mathcal{F}_1)$ is solved quickly, lower bounds can alternatively be computed 
by applying B\&B to $\mathcal{F}_1$ for a restricted amount of time.
While the initial bound $z\left(\text{LP}(\mathcal{F}_1)\right)$ might be weak, fast solution of $\text{LP}(\mathcal{F}_1)$ enables rapid growth in the B\&B-tree,
and, possibly, also in the corresponding lower bound.

@@

    \begin{subfigure}[b]{0.49\textwidth}
        \includegraphics[width=\textwidth]{lower-bound-32-16}
        \caption{$|V|=32, |D|=16$}
        \label{fig:cggr32-16}
    \end{subfigure}
    \hfill %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc. 
      %(or a blank line to force the subfigure onto a new line)
    \begin{subfigure}[b]{0.49\textwidth}
        \includegraphics[width=\textwidth]{lower-bound-34-17}
        \caption{$|V|=34, |D|=17$}
        \label{fig:cggr34-17}
    \end{subfigure}
  \caption{Comparison of lower bounds obtained by CG and B\&B on model $\mathcal{F}_1$.
	\textcolor{blue}{Each figure depicts a progress of lower bounds for instances with varying number of destination and non-destination nodes.
	The lower bounds are obtained by solving CG scheme applied on model $\mathcal{F}_1$ (black curves), and produced during the course of classic B\&B algorithm (grey curves)
	The tightness of the bounds increases with time, and it can be observed from the graphs that the CG bound becomes tighter at some point.
	We express the quality of lower bounds with respect to the value of simple continuous relaxation, as indicated by the vertical axes.
	It can also be noted that the lower bound obtained from B\&B grows sharply during the computation of the simplex method in the root of the B\&B tree, 
	which is happening until the curve reaches 100\%.
	After that point the bound grows rather slowly.
	CG on the other hand produces bound that grows faster even after the value of first relaxation in the root is reached.
%	The values can be retrieved from the output of CPLEX during the course of simplex and B\&B algorithm.
	}}
  \label{fig:cggr}
\end{figure}

>>>>

The practical part of this work focuses on comparison of models presented in Sections \ref{sec:ILP} and \ref{sec:strong}.
Instances of given number of nodes and destinations are generated with random coordinates uniformly distributed on the square with corners at [0, 0] and [100, 100].
A time limit of 20 CPU-minutes is imposed on each run.
All experiments are run on an Intel Core 2 Quad CPU at 2.83 GHz and 8 GB RAM.
The models are implemented in Java with Concert technology and solved using Cplex 12.5.1 optimizer.

@@

Small instances are often solved to optimality relatively quickly with B\&B.
We investigate what is the maximum size of instances that can be solved to optimality within 20 CPU-minutes using models $\mathcal{X}_1$ and $\mathcal{F}_1$.
The experiments compare sets of instances with different $|V|/|D|$ ratios.
For instances of the size given in columns 1--2, Table \ref{tab:small_inst} shows the average running time in CPU-seconds (columns 3--4),
the number of instances solved to optimality (columns 5--6),
and the average ratio between the upper and lower bounds on the minimum cost (columns 7--8) upon interruption of the solver.
Each table value is obtained after solving 25 instances of given size.

@@

\textcolor{blue}{\subsection{Comparison with alternative methods}}
\textcolor{blue}{
We now investigate the ability of IP models $\mathcal{X}_1$ and $\mathcal{F}_1$ to produce a feasible solution when a time limit is imposed.
We compare the quality of obtained solutions with those calculated by Alg. \ref{alg:meta}.
The results for time restrictions 5, 10 and 20 CPU-minutes are reported in Tab. \ref{tab:metamod}.
These experiments are carried out on the same instance set that is used in experiments presented in Tab. \ref{tab:small_inst}. 
}
\begin{table}[]
\centering
\setlength{\tabcolsep}{6pt} % Default value: 6pt
\renewcommand{\arraystretch}{1.4} % Default value: 1
\textcolor{blue}{
\begin{tabular}{rrrrrrrrrrr}
\multicolumn{1}{l}{}      & \multicolumn{1}{l}{}      & \multicolumn{3}{c}{5 min}  & \multicolumn{3}{c}{10 min} & \multicolumn{3}{c}{20 min}     \\\hline
\multicolumn{1}{l}{$|V|$} & \multicolumn{1}{l}{$|D|$} & \multicolumn{1}{c}{$\mathcal{X}_1$} & \multicolumn{1}{c}{$\mathcal{F}_1$} & \multicolumn{1}{c}{$\mathcal{M}$} & \multicolumn{1}{c}{$\mathcal{X}_1$} & \multicolumn{1}{c}{$\mathcal{F}_1$} & \multicolumn{1}{c}{$\mathcal{M}$} & \multicolumn{1}{c}{$\mathcal{X}_1$} & \multicolumn{1}{c}{$\mathcal{F}_1$} & \multicolumn{1}{c}{$\mathcal{M}$} \\
\hline
18  & 12   &100.1&100.1&100.0&100.0&100.0&100.0&100.0&100.0&100.0\\
21  & 14   &107.9&108.8&100.0&102.7&103.2&100.0&101.2&100.7&100.0\\
24  & 16   &153.4&115.6&100.0&126.2&114.2&100.0&104.3&112.7&100.0\\     
\hline
20  & 10   &100.6&100.1&100.0&100.1&100.0&100.0&100.1&100.0&100.0\\
22  & 11   &102.8&103.9&100.0&101.3&100.7&100.1&100.7&100.2&100.1\\
24  & 12   &108.2&113.4&100.0&105.0&105.5&100.0&103.1&106.2&100.0\\                               
\hline
21  & 7    &100.1&100.0&100.0&100.0&100.0&100.0&100.0&100.0&100.0\\
24  & 8    &102.8&101.9&100.0&102.1&100.7&100.0&101.1&100.1&100.0\\
27  & 9    &133.2&118.7&100.0&111.3&109.1&100.0&108.2&104.1&100.0                                  
\end{tabular}
}
\caption{\textcolor{blue}{Comparison of the metaheuristic algorithm and solving IP models with imposed time-limit.}}
\label{tab:metamod}
\end{table}

\textcolor{blue}{
Each entry contains the average percentage  of the objective value yielded by the most successful method in a given instance type.
For example, if all methods find the optimum in every instance, the three corresponding values are all 100.0.
Thus, the lowest value in a given instance type indicates that the corresponding method performs the most successfully. 
The averages are calculated over all instances of given $|V|$ and $|D|$ (25 instances for each type).
}

\textcolor{blue}{
The resulting values reveal that the metaheuristic algorithm converges quickly, and is often able to find optimal solutions faster than Cplex applied to the IP models.
The smallest 5 CPU-minutes time limit suffices to find a good quality solutions in our test scenarios, and increasing the time limit rarely improves the outcome.
The solutions are, however, not guaranteed to be optimal.
The optimality is proven only for smaller instances, where the B\&B finds the optimum before the time limit is exceeded.
It is likely that the inexact method executed on large instances fails to find the optimum, but it very clearly outperforms B\&B applied to the IP models, 
where the solutions found within the given time limit have far larger objective function value.
}
\newline

@@

As the instance size increases, exact solution of the SMT problem becomes unrealistic.
To assess the quality of fast and possible inexact methods, \textcolor{blue}{such as metaheuristic Alg. \ref{alg:meta}, 
or construction methods by \citep{ivanova16isco}},  access to tightest possible lower bounds is crucial.
In the next set of experiments, we compare two different methods for computing the bounds.
Section \ref{sec:expcomplp} indicates that $z\left(\text{LP}(\mathcal{X}_3)\right)$ is a strong bound,
but solving $\text{LP}(\mathcal{X}_3)$ requires excessive computations.
At this point, the CG scheme (Alg.\ \ref{alg:cg}) comes into play.
Because $\text{LP}(\mathcal{F}_1)$ is solved quickly, lower bounds can alternatively be computed 
by applying B\&B to $\mathcal{F}_1$ for a restricted amount of time.
While the initial bound $z\left(\text{LP}(\mathcal{F}_1)\right)$ might be weak, fast solution of $\text{LP}(\mathcal{F}_1)$ enables rapid growth in the B\&B-tree,
and, possibly, also in the corresponding lower bound.

@@

    \begin{subfigure}[b]{0.49\textwidth}
        \includegraphics[width=\textwidth]{lower-bound-32-16}
        \caption{$|V|=32, |D|=16$}
        \label{fig:cggr32-16}
    \end{subfigure}
    \hfill %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc. 
      %(or a blank line to force the subfigure onto a new line)
    \begin{subfigure}[b]{0.49\textwidth}
        \includegraphics[width=\textwidth]{lower-bound-34-17}
        \caption{$|V|=34, |D|=17$}
        \label{fig:cggr34-17}
    \end{subfigure}
  \caption{Comparison of lower bounds obtained by CG and B\&B on model $\mathcal{F}_1$.
	\textcolor{blue}{Each figure depicts the progress of lower bounds for instances with varying number of destination and non-destination nodes.
	The lower bounds are obtained by solving CG scheme applied on model $\mathcal{F}_1$ (black curves), and produced during the course of classic B\&B algorithm (grey curves)
	The tightness of the bounds increases with time, and it can be observed from the graphs that the CG bound becomes tighter at some point.
	We express the quality of lower bounds with respect to the value of simple continuous relaxation, as indicated by the vertical axes.
	It can also be noted that the lower bound obtained from B\&B grows sharply during the computation of the simplex method in the root of the B\&B tree, 
	which is happening until the curve reaches 100\%.
	After that point, the bound grows rather slowly.
	CG on the other hand produces bound that grows faster even after the value of first relaxation in the root is reached.
%	The values can be retrieved from the output of CPLEX during the course of simplex and B\&B algorithm.
	}}
  \label{fig:cggr}
\end{figure}
